---
title: "Part II Project - Illumina Analysis"
author: "Dalia Bornstein"
date: "2024-11-21"
output:
  github_document:
    toc: true
    fig_width: 8
    fig_height: 8
  pdf_document:
    fig_width: 7
    fig_height: 6
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Libraries
```{r message=FALSE}
library(RColorBrewer); packageVersion("RColorBrewer")
library(dada2); packageVersion("dada2")
library(kableExtra) ; packageVersion("kableExtra")
library(microbiome); packageVersion("microbiome")
library(microViz); packageVersion("microViz")
library(phyloseq); packageVersion("phyloseq")
library(Biostrings); packageVersion("Biostrings")
library(ggplot2); packageVersion("ggplot2")
theme_set(theme_bw()); # Set ggplot theme
library(dplyr); packageVersion("dplyr")
library(gdata); packageVersion("gdata")
```

```{r}
path <- "rawReads_16s"
list.files(path)
```

# List Names in Functional Way
```{r}
# Forward and reverse fastq filenames have format

fnFs <- sort(list.files(path, pattern="_R1.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2.fastq.gz", full.names = TRUE))

# Extract sample names, assuming filenames have format

sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)

kable(sample.names,caption = "Sample Names")
```


# Plot Quality Profiles

## Forward
```{r fig.width=10,fig.height=7,dpi=300}
plotQualityProfile(fnFs)
```

## Reverse

Look at quality for the second read of each pair in the aggregate.
```{r fig.width=10,fig.height=7,dpi=300}
plotQualityProfile(fnRs)
```


# Filter and Trim

We want to filter out reads that look bad already and to filter reads that are too short etc. Additionally we will trim the left read at 221nt and the right at 141nt to remove the most error-prone regions. We will remove PhiX contamination at the same time.

```{r}
# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(221,141),
              maxN=0, truncQ=2, rm.phix=TRUE,
              compress=TRUE, multithread=TRUE) 
```

```{r}
kable(out)
```
# Learn Error Rates

Statistical modelling of error rates from a sample subset. Should be then applied over all datasets.
```{r}
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
```

# Plot Error Rates

Look at the different types of errors that occur against our model of Illumina Error Rates.
```{r fig.width=10,fig.height=7, dpi=300}
plotErrors(errF, nominalQ=TRUE)
```

# Sample Inference

We aim to identify ASVs (Amplicon Sequence Variants from our data). This is a unique DNA sequence that is generated from a targeted amplification of specific regions in a microbial genome. These tend to be a better surrogate of microbial abundance than an OTU (Operational Taxonomic Unit). This is DADA2s standard and best approach.

Forward Reads first.
```{r}
dadaFs <- dada(filtFs, err=errF, multithread=TRUE,pool=FALSE)
```

Now Reverse Reads (2nd of Pair).
```{r}
dadaRs <- dada(filtRs, err=errR, multithread=TRUE,pool=FALSE)
```
## Results of Denoising
```{r}
dadaFs[[1]]
```


# Merge and Construct Table

Lets merge our forward and reverse reads into a single amplicon sequence construct. This is done by concatenation as our reads do not overlap.

We will pad the space with an appropriate number of `N` nucleotides. This causes some issue later for species assignment that we will deal with.

```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE,justConcatenate=TRUE)
mergers2 <- mergers
 
# Repad to 28nt of Ns for an insert of 490ish
for (i in 1:length(names(mergers))){
  #  print(names(mergers)[i])
  mergers2[[i]]$sequence=gsub("NNNNNNNNNN","NNNNNNNNNNNNNNNNNNNNNNNNNNNN",mergers[[i]]$sequence)
}
 
seqtab <- makeSequenceTable(mergers2)
dim(seqtab)
table(nchar(getSequences(seqtab)))
```


# Remove Chimeras

Identify and remove PCR Chimeras. These are separate fragments that somehow form a chimeric pair during PCR.

This is usually when an aborted PCR fragment forms a primer for another sequence.
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
```

## Fraction of Chimeric Reads per sample

Lets compute how many chimeras we actually had per sample.
```{r}
1-sum(seqtab.nochim)/sum(seqtab)
```


# Track Reads through Pipeline
```{r fig.width=10,fig.height=7, dpi=300}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
kable(track,caption = "Sample Filtering Table")

colours=brewer.pal(n = 10, name = "Set3")
barplot(track[,1],las=2,cex.names=0.5,cex.axis=0.6,col=colours,main="Raw Illumina Readpairs per Sample")

pie(track[,1]/(sum(track[,1])),labels = paste0(names(track[,1]),"\n",as.numeric(track[,1]/(sum(track[,1])))*100,"%"),cex=0.5,col=colours,main="Raw Illumina Readpairs per Sample")
 
colours=brewer.pal(n = 6, name = "Set1")
names(colours)=c("input","filtered","denoisedF","denoisedR","merged","nonchim")
barplot(t(track),beside=TRUE,las=2,cex.axis=0.5,cex.names=0.5,col=colours,main="Read Level Filtering")
legend("bottomright",names(colours),fill=colours,cex=0.4)
```

# Taxonomic Assignment (SILVA)

We use version 132 of the SILVA 16s rRNA database for bacteria to assign. We use the training set for standard assignment and a separate species level file for species level assignment where possible.
```{r}
# Assign Taxonomy
silva_train=paste0("./16s_taxonomy/silva_nr_v132_train_set.fa.gz")
silva_species=paste0("./16s_taxonomy/silva_species_assignment_v132.fa.gz")
 
taxa <- assignTaxonomy(seqtab.nochim, silva_train, multithread=TRUE,verbose=TRUE)
```

## Wrangling Species Calls

We have 'N's in our sequences, so we cant just ask for species calls as DADA2 doesn't like Ns. Instead we will make two new taxa objects.
taxa2 will remove all the N's and everything to the right of the junction and try and assign species using only the left-most sequence.
taxa3 will remove all the N's and everything to the left of the junction and try and assign species using only the right-most sequence.
Between the left and right side analysis we should be able to get at least some species level calls for our dataset.
```{r}
taxa2 <- taxa
rownames(taxa2)=gsub("NNNNNNNNNN.*$","",rownames(taxa2))
taxa2 <- addSpecies(taxa2, silva_species,verbose=TRUE)
 
taxa3 <- taxa
rownames(taxa3)=gsub("^.*NNNNNNNNNN","",rownames(taxa3))
taxa3 <- addSpecies(taxa3, silva_species,verbose=TRUE)
```

We now will make a new species column in our original taxa object to add this species data to, we'll first populate it with the species calls from taxa2.
```{r}
if (ncol(taxa)==6){
taxa=cbind(taxa,taxa2[,7])
colnames(taxa)[7]="Species"
taxa[,7]=NA;
}
```

## Merge Species Back to taxa

Now we will add the extra species data from taxa3. Once done we should have everything we need in the taxa object and no longer need taxa2 and taxa3.
```{r}
for (i in 1:nrow(taxa)){
 if ((!is.na(taxa2[i,7])) && (!is.na(taxa3[i,7]))){
  taxa[i,7]=taxa2[i,7]
 } else {
   if (!is.na(taxa2[i,7])){
    taxa[i,7]=taxa2[i,7]
   }
   if (!is.na(taxa3[i,7])){
    taxa[i,7]=taxa3[i,7]
   }
 }
}

rm(taxa2)
rm(taxa3)
```

# Save Taxonomic Objects
We could decide to save the R objects now, so that they can be reloaded later.
We could then load these in the next analysis script as needed.
```{r}
#saveRDS(seqtab.nochim,file="seqtab.rds")
#saveRDS(taxa2,file="taxa2.rds")
#saveRDS(taxa,file="taxa.rds")
#saveRDS(taxa3,file="taxa3.rds")
```

## Illumina Microbiome Analysis

Now we move to *phyloseq* and use multiple plots to explore microbiome prevalence and abundance across samples. 

## Prepare Object Table 

Create some simple tables including samdf for our samples, display our sample table here.
```{r}
samples.out <- rownames(seqtab)
subject <- samples.out
samdf <- data.frame(Subject=subject)
rownames(samdf) <- samples.out
```

## Make our phyloseq object for subsequent analysis.
```{r}
ps <- phyloseq(otu_table(seqtab, taxa_are_rows=FALSE), 
               sample_data(samdf), 
               tax_table(taxa))
```

## Analysis of ASV assignment across taxonomy per sample

Lets explore how each sample did with taxonomic assignment across the levels.
```{r fig.width=10,fig.height=7, dpi=300}
ps.melt <- ps %>% psmelt()
taxLevels = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

table=data.frame(samples.out)
rownames(table)=table$samples.out

for (i in 1:length(taxLevels)){
  table[,taxLevels[i]]=0
  for (j in 1:length(samples.out)){
    subt= ps.melt[ps.melt$Sample==samples.out[j],]
    table[samples.out[j],taxLevels[i]]=sum(subt[!is.na(subt[,taxLevels[i]]),"Abundance"])
  }
}

colours=brewer.pal(n = 7, name = "Set2")
names(colours)=taxLevels
barplot(t(table[,2:8]),beside=TRUE,las=2,cex.axis=0.5,cex.names=0.5,col=colours,main="Resolution to Taxonomic Level",sub="Abundance Per sample")
legend("topright",names(colours),fill=colours,cex=0.4)

```

## Process the DNA data also for possible phylogenetics
```{r}
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps
```
# Analysis

## Visualise Alpha-Diversity

Lets look at microbial diversity and richness across the sample space.

```{r fig.width=10,fig.height=7, dpi=300}
sample_data(ps)$SampleID <- rownames(sample_data(ps))
plot_richness(ps, x="SampleID", measures=c("Shannon", "Simpson"), color="SampleID")
```


## Ordinate Analysis

```{r fig.width=10,fig.height=7, dpi=300}
# Transform data to proportions as appropriate for Bray-Curtis distances
ps.prop <- transform_sample_counts(ps, function(otu) otu/sum(otu))
ord.nmds.bray <- ordinate(ps.prop, method="NMDS", distance="bray")

plot_ordination(ps.prop, ord.nmds.bray, color="SampleID", title="Bray NMDS")

top20 <- names(sort(taxa_sums(ps), decreasing=TRUE))[1:20]
ps.top20 <- transform_sample_counts(ps, function(OTU) OTU/sum(OTU))
ps.top20 <- prune_taxa(top20, ps.top20)
plot_bar(ps.top20, x="SampleID", fill="Family") + facet_wrap(~Subject, scales="free_x")
plot_bar(ps.top20, x="SampleID", fill="Genus") + facet_wrap(~Subject, scales="free_x")

```

## Relative Abundance by Family

```{r fig.width=10,fig.height=7, dpi=300}
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
ps_filtered <- filter_taxa(ps_rel, function(x) mean(x) > 0.0001, TRUE)
plot_bar(ps_filtered, x = "SampleID", fill = "Family") +
  theme_minimal() +
  labs(y = "Relative Abundance")
```

## Prevalence Analysis by Phyla

Lets now compute new Prevlance Tables by Phylum and then Genus.

```{r fig.width=10,fig.height=7}
prevdf <- apply(X = otu_table(ps),
MARGIN = ifelse(taxa_are_rows(ps), yes = 1, no = 2),
FUN = function(x){sum(x > 0)})
# Add taxonomy and total read counts to this data.frame
prevdf <- data.frame(Prevalence = prevdf, TotalAbundance = taxa_sums(ps),
tax_table(ps))
ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(ps),color=Phylum)) +
  # Include a guess for parameter
geom_point(size = 2, alpha = 0.7) +
geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) +
scale_x_log10() + xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") + facet_wrap(~Phylum) + theme(legend.position="none") + ggtitle("Bacteria Prevalence Phylum")
```
## Prevalence Analysis by Genus

Same again, but now by Genus level. Perhaps too much detail.

```{r fig.width=10,fig.height=12}
prevdf <- apply(X = otu_table(ps),
MARGIN = ifelse(taxa_are_rows(ps), yes = 1, no = 2),
FUN = function(x){sum(x > 0)})
# Add taxonomy and total read counts to this data.frame
prevdf <- data.frame(Prevalence = prevdf, TotalAbundance = taxa_sums(ps),
tax_table(ps))
ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(ps),color=Genus)) +
  # Include a guess for parameter
geom_point(size = 1, alpha = 0.7) +
geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) +
theme(strip.text = element_text(size = rel(0.5))) +
scale_x_log10() + xlab("Total Abundance") + ylab("Prevalence [Frac. Samples]") + facet_wrap(~Genus) + theme(legend.position="none") + ggtitle("Bacteria Prevalence Genus")
```

## Nicer Plots for Diversity and Richness

```{r fig.width=10,fig.height=7}


colour_c=brewer.pal(n = length(samdf$Subject), name = "Set3")
names(colour_c)=samdf$Subject

p = plot_richness(ps, x="Subject", title="Diversity and Richness", color="Subject", measures=c("Shannon","Simpson")) 
p + geom_point(size=5, alpha=0.7) + scale_color_manual(values=colour_c)

```

## Ordination Plot (PCA like) Bray Curtis Distance

```{r fig.width=10,fig.height=7}
ps.ord <- ordinate(ps, "PCoA", "bray")
p=plot_ordination(ps, ps.ord, type="samples", label = "Subject", color = "Subject", title = "Bacterial Bray-Curtis Distance by Sample Type") 
p + geom_point(size=5, alpha=0.7) + scale_color_manual(values=colour_c)
```


## Prettier Genus Level Plots

```{r fig.width=10,fig.height=7}
ps16=ps
psq <- ps16 %>%
  tax_filter(min_prevalence = 5) %>%
  tax_fix()
psq %>%
  comp_barplot(tax_level = "Genus",n_taxa=20,label="Subject") +
  coord_flip() + theme(legend.key.size = unit(0.2, 'cm'),legend.text = element_text(size=5))

psq %>%
  comp_barplot(tax_level = "Genus",n_taxa=20,label="Subject") + 
  theme(legend.key.size = unit(0.2, 'cm'),
        legend.text = element_text(size=5),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size=5))
```


## Prettier Genus and Species Plots

```{r fig.width=10,fig.height=7}

psq %>%
  comp_barplot(tax_level = "Order",n_taxa=10,label="Subject") +
  coord_flip() + theme(legend.key.size = unit(0.2, 'cm'),legend.text = element_text(size=5))

psq %>%
  comp_barplot(tax_level = "Order",n_taxa=10,label="Subject") +
  theme(legend.key.size = unit(0.2, 'cm'),
        legend.text = element_text(size=5),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size=5))

psq %>%
  comp_barplot(tax_level = "Species",n_taxa=20,label="Subject") +
  coord_flip() + theme(legend.key.size = unit(0.2, 'cm'),legend.text = element_text(size=5))

psq %>%
  comp_barplot(tax_level = "Species",n_taxa=20,label="Subject") +
  theme(legend.key.size = unit(0.2, 'cm'),
        legend.text = element_text(size=5),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size=5))

```

```{r,fig.width=10,fig.height=7}

psq %>%
  comp_barplot(tax_level = "Species",n_taxa=20,label="Subject") +
  coord_flip() + theme(legend.key.size = unit(0.2, 'cm'),legend.text = element_text(size=5))

psq %>%
  comp_barplot(tax_level = "Species",n_taxa=20,label="Subject") +
  theme(legend.key.size = unit(0.2, 'cm'),
        legend.text = element_text(size=5),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size=5))

```

## Iris Plot of Genus Level Ordination

```{r,fig.width=10,fig.height=7}
psq %>%
  tax_transform("clr", rank = "Genus") %>%
  # when no distance matrix or constraints are supplied, PCA is the default/auto ordination method
  ord_calc() %>%
  ord_plot_iris(tax_level = "Genus", ord_plot = "none", n_taxa = 20, anno_colour = "Subject") +
  theme(legend.key.size = unit(0.2, 'cm'),legend.text = element_text(size=5))
```

## Iris Plot of Species Level Ordination

```{r,fig.width=10,fig.height=7}
psq %>%
  tax_transform("clr", rank = "Species") %>%
  # when no distance matrix or constraints are supplied, PCA is the default/auto ordination method
  ord_calc() %>%
  ord_plot_iris(tax_level = "Species", ord_plot = "none", n_taxa = 20, anno_colour = "Subject") +
  theme(legend.key.size = unit(0.2, 'cm'),legend.text = element_text(size=5))
```



