---
title: "Part II Project - Nanopore Count Analysis"
author: "Dalia Bornstein"
date: "2024-11-21"
output:
  github_document:
    toc: true
    fig_width: 8
    fig_height: 8
  pdf_document:
    fig_width: 7
    fig_height: 6
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# R Analysis of Nanopore Taxonomic Count Summaryies.

These have been produced from the Kraken Report files using a perl script [attached](scripts/read_counts.pl)

## Load Libraries

```{r message=FALSE}
library(RColorBrewer); packageVersion("RColorBrewer")
library(kableExtra) ; packageVersion("kableExtra")
library(ggplot2); packageVersion("ggplot2")
theme_set(theme_bw()); # Set ggplot theme
library(dplyr); packageVersion("dplyr")
library(gtools); packageVersion("gtools")
```

## Read the Table

Read in [nanopore_counts.txt](tables/nanopore_counts.txt) and make a bar-chart. The first numeric column is unmapped reads, the remaining columns decay from the total number of mapped reads.
We will reorder the table as the columns are not sorted alphanumerically. The Nanopore_counts.txt table was made by the ancillary script [read_counts.pl](scripts/read_counts.pl) on the kraken files as follows:

`perl read_counts.pl barcode*.krakenreport.txt > nanopore_counts.txt`

```{r}
nanopore_table <- read.table("tables/nanopore_counts.txt",header=TRUE,row.names=1,sep="\t")
subtable=nanopore_table[,c(2:11)]
subtable=subtable[,mixedorder(colnames(subtable))]
nanopore_table[,c(2:11)] <- subtable
colnames(nanopore_table)[2:11] <- colnames(subtable)
kable(nanopore_table[c(1,2,4,6,7,8,9,10,11,12,13),],caption = "Nanopore Metagenomic Classification Levels per Sample")

barplot(as.matrix(nanopore_table[c(1,2,4,6,7,8,9,10,11,12,13),c(2:11)]),beside=T,col=rainbow(11),las=1,names=c(1:10),ylab="No. of Reads")
legend("topright",nanopore_table[c(1,2,4,6,7,8,9,10,11,12,13),1],fill=rainbow(11))
```

## PDF Output

We will save a PDF version of this plot in the pdf folder for simplicity.
```{r}
pdf("pdf/nanopore_tax_levels.pdf",paper="a4r")
barplot(as.matrix(nanopore_table[c(1,2,4,6,7,8,9,10,11,12,13),c(2:11)]),beside=T,col=rainbow(11),las=1,names=c(1:10),ylab="No. of Reads")
legend("topright",nanopore_table[c(1,2,4,6,7,8,9,10,11,12,13),1],fill=rainbow(11))
dev.off()
```
